---
title: "2029 U.S. Job Market Evaluation Proposal"
author: "Roger Henderson"
output:
  pdf_document: default
  #html_document: default
---

 
```{r echo=FALSE}
  monthname <- months(Sys.Date()) 
 
  dayyr <- format(Sys.Date(), "%d, %Y")

  paste("Date: ", monthname, dayyr, sep = " ")
```

### A.)	Project Request:
        Ten years from now, in 2029, there are likely to be significant changes in the make-up of the 
        United States workforce. Two unique questions come to mind:
        
        1.) Approximately what percentage of the United States workforce will be working in the gig 
            economy? The gig economy is defined as work that is temporary and flexible.
        2.)	What percentages will be working remotely, in contract, freelance, or in some other form of a 
            non-traditional on-site position?
            
        The goal of this project is to gather an understanding of the existing market and trends to 
        forecast out employee makeup ten years into the future. The project is centered around core Amazon 
        business units; not the actions of subsidiaries.
        
### B.)	Stakeholder(s):

        The primary client is Beth Galetti, Amazon's Senior Executive VP of Human Resources [7]. She is 
        primarily responsible for the planning of facilities and workspaces needed as part of the HQ2 
        expansion. The Executive VP wants to make effective use of resources and minimize expenses while 
        ensuring there are adequate workspaces available. Regardless of the Amazon timeline to begin 
        operations in the HQ2 facility, a long-term study is needed to effectively plan for the new 
        facility and the first few years of operational staffing. 

        The list of U.S. finalist cities as of the end of 2017 in alphabetical order are: Atlanta, Austin, 
        Boston, Chicago, Columbus, Dallas, Denver, Indianapolis, Los Angeles, Miami, Montgomery County 
        (PA), Maryland, Nashville, Newark, New York, Northern Virginia, Philadelphia, Pittsburgh, Raleigh, 
        and Washington D.C. Please note the shortlist is primarily centered around primarily east coast 
        cities [4].

        If considerably more talented workers will be found in the gig economy within the next ten years, 
        dramatic changes will need to be made in HR policies. Some of these changes include flexible       
        short-term compensation and benefits packages, an increase in wireless connectivity and security
        options and infrastructure to support remote operations, tax concerns related to free-lance     
        employees, the redefinition of a team and how to achieve team cohesion, etc.

        Decisions will need to be made at geographical, work classification (gig, seasonal temporary, 
        contract, full-time permanent), and department (fulfillment center, driver, administrative, 
        software development, analytics) levels.

        Per Amazon's 2017 Annual Report on the Investor Relations site, as of December 31, 2017, the 
        company employed approximately 566K full-time and part-time employees. In 2017 alone, Amazon 
        directly created more than 130K new permanent, full-time Amazon jobs, not including acquisitions, 
        bringing its permanent global employee base to just over 560K [1].

        The Company also utilizes independent contractors and temporary personnel to supplement the 
        workforce in three main channels. The first channel is utilizing temporary seasonal workers in its 
        warehouses and customer service centers. According to CNET and Fortune Magazine, utilizing 2016 
        data, Amazon traditionally hires between 70K to 120K seasonal employees, with more than 14K of 
        those positions converting to regular, full-time roles [5] [13]. Considerations for seasonal 
        workers center on providing equitable contracts and ensuring available work center capacity.

        There is limited information available to the public however it is important to recognize that most
        of Amazon's non-traditional workers are temporary seasonal workers and the number of season workers
        increases 5K to 10K each year [5]. Fulfillment and customer care centers can currently be found in 
        the U.S. regions listed below.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=FALSE}

# Definition of vectors
region <- c("New England:", "Atlantic:", "Southeast:", "Central:", "Midwest:", "Southwest:",
            "Mountain:", "Pacific:")

state <- c("Connecticut, Delaware, Massachusetts, New Hampshire",
           "Maryland, New Jersey, New York, North Carolina, etc.",
           "Alabama, Florida, Georgia, South Carolina",
           "Illinois, Indiana, Kentucky, Michigan, Ohio, etc.",
           "Kansas, Minnesota, Oklahoma",
           "Texas",
           "Arizona, Colorado, Nevada, Utah",
           "California, Oregon, Washington")

# Create a data frame from the vectors
fullfillment_df <- data.frame(region, state)

fullfillment_df


```



```{r}

```

        The second channel of temporary workers is Amazon Flex gig drivers who deliver Amazon Prime 
        packages to customers. Glassdoor estimates the number of these flexible employees are between 
        5K to 10K employees [6]. Considerations for gig workers center on effective training and 
        clearly defined roles to ensure all local, state, and federal tax and gig employment guidelines 
        are being followed. Amazon would not wish to incorrectly classify contract workers as gig employees.

        Amazon Flex is located across the U.S with operations located in New England, Atlantic, Southeast, 
        Central, Midwest, Southwest, Mountain, and Pacific regions. Below is the list of cities in which 
        Amazon Flex is offered with the corresponding U.S. census geographical region [15]. 


```{r echo=FALSE}

# Definition of vectors
region <- c("New England:", "Atlantic:", "Southeast:", "Central:", "Midwest:", "Southwest:",
            "Mountain:", "Pacific:")

state <- c("Boston",
           "New Jersey, Philadelphia, Pittsburgh, Charlotte",
           "Miami, Orlando, Tampa",
           "Detroit, Milwaukee, Minneapolis",
           "Kansas City, St. Louis, Kentucky Area",
           "Dallas",
           "Denver, Tucson",
           "Pacific - Greater Los Angeles Area, etc.")

# Create a data frame from the vectors
amz_flex_df <- data.frame(region, state)

amz_flex_df


```

        As this is a relatively new service, there is not enough data currently available to estimate Amazon 
        Flex specific growth trends. However, we can surmise that the service is likely to expand further 
        because after Amazon's acquisition of Whole Foods in August 2017, Amazon Flex has steadily expanded 
        into new locations [12].
        
        The third channel of temporary workers is personnel in the corporate office who are employed through 
        staffing agencies. Contractors working at Seattle, Washington headquarters locations are comprised of 
        the following occupations and approved contract agencies:
        
```{r echo=FALSE}

# Definition of vectors
job_functions <- c("IT, Web, & Software", "Finance & Accounting", "Administrative Functions", "Design & Creative", "Manufacturing", "Sales & Marketing","HR & Recruiting", "Engineering")

staffing_agencies <- c("OSI Engineering, Prime 8 Consulting, etc.",
           "Affirma Consulting, Robert Half, Rylem, etc.",
           "Aerotek, Concentrix Daksh, Corestaff, etc.",
           "Creative Circle, Filter, Simplicity Consulting",
           "Aerotek, Integrity Staffing Solutions, etc.",
           "Kforce, Simplicity Consulting",
           "Rylem, Simplicity Consulting",
           "OSI Engineering")

# Create a data frame from the vectors
amz_contract_df <- data.frame(job_functions, staffing_agencies)

amz_contract_df


```     

        Amazon does not release the number of corporate contract employees. The company has U.S. software 
        development centers in the locations listed below [2-3, 8-11, 14].      
        
                
```{r echo=FALSE}

# Definition of vectors
region <- c("New England:", "Atlantic:", "Midwest:", "Southwest:", "Mountain:", "Pacific:")

city <- c("Cambridge",
           "New York, Pittsburgh, Herndon (VA)",
           "Minneapolis",
           "Austin",
           "Temple",
           "Cupertino, Irvine, San Francisco, San Luis Obispo, Seattle")

# Create a data frame from the vectors
amz_sfwt_dev_df <- data.frame(region, city)

amz_sfwt_dev_df


```  

        Amazon may wish to recruit more contract employees because there is tenure flexibility based on 
        production needs and the company can reduce risk by getting an opportunity to try out an employee 
        to determine if they are a good fit before committing to hiring them. 

        Staffing agencies usually help organizations cut expensive hiring costs and create a candidate 
        profile that engender better job candidates. There are legal issues to consider when considering 
        staffing company employees including the possible misclassification of employees for tax purposes, 
        compliance with labor law and federal laws, and possible contract disputes. 

        In addition to the considerations that arise with gig and contract workers, if more permanent future 
        employees expect to have partial or full remote working capability in HQ2 facilities this has a 
        significant impact on the extent of the construction. 

        If more employees will wish to work remotely, Amazon will want to accommodate the needs of the workforce 
        and can save considerable expense in construction, rent, and utilities. Amazon can utilize the findings 
        of this analysis to better decide how many facilities and the dimensions of the facilities needed to 
        accommodate its onsite HQ2 personnel whether they be contract, gig, or permanent. 

        In addition to the construction and ongoing expenses related to urban workspaces, Amazon may wish to lead 
        the effort to employ more gig and remote workers, to obtain other benefits. These benefits include access 
        to a bigger talent pool not limited by geographical location, tapping into less inflated high salary job 
        markets, and accessing remote workers who are generally happier and more productive. 

        If there is a significant trend to more flexible working options, HR will want to begin long-range planning. 
        The planning could include hiring and training leaders that can effectively manage virtually with the right 
        balance of autonomy and accountability, determining the best options for setting up high capacity remote 
        systems connectivity, legal employee contact changes, and several other considerations.

### C.)	Data Sources:

       . Department of Labor: Closely watched measures of employment and unemployment. https://www.dol.gov/
       
       . Employment by U.S. Census: Data that measures the state of the nation's workforce, including employment
         and unemployment levels, as well as weeks and hours worked. https://www.census.gov/topics/employment.html
         
       . US Government Open Data: United States open source software government site making data open and 
         machine-readable on a variety of topics. https://www.data.gov/open-gov/
         
       . Bureau of Labor Statistics: U.S. government's data collection of employment-related stats across 
         regions, states, and local areas. 
         https://www.bls.gov/opub/mlr/2018/article/electronically-mediated-work-new-questions-in-the-contingent
         -worker-supplement.htm
         
       . JP Morgan Chase Online Platform Economy in 2018: JP Morgan Chase's evaluation of the global economy 
         which provides gig economy findings https://www.jpmorganchase.com/corporate/institute/document/
         institute-ope-2018.pdf
         
       . MIT Technology Review: MIT online article that addresses how the US Bureau of Labor Statistics is 
         likely undercounting alternative work arrangement employees.      
         https://www.technologyreview.com/s/611381/the-us-government-is-seriously-underestimating-how-much-
         americans-rely-on-gig-work/

### D.)	Approach:
         
        The general approach to answering the 2029 United States gig and remote workers percentages questions 
        are listed below.

        1. Obtain data on current gig and remote workers makeup. The goal is to gain a U.S employee 
           multi-dimensional data set that includes the job classification, demographic factors, 
           geographic factors, economic factors, plus the fraction of gig/remote workers.  
        2. Gain an understanding if the trend is increasing or decreasing and by what percentages 
           in the U.S.
            a.	Employ the statistical classification model to classify current employment data points 
                into increases or decreases in gig and remote employment. 
            b.	Employ the clustering classification model to group like employment data points into 
                increases or decreases in gig and remote employment clustered by geographical region. 
            c.	Employ multiple regression analysis models to estimate the relationship between employment 
                data points to determine an upward or downward trend in gig and remote employment.
        3. If all three models (classification, clustering, and regression show a consistent upward or 
           downward trend) gain an understanding of the factors behind the trend.
        4. Adjust the data points, using the factors behind the trend, to obtain new 2029 data points. 
        5. Rank the potential U.S. future employment scenarios using cluster analysis.
        6. Apply trend analysis to update the estimates to get the final estimated impact by job type on 
           the Amazon HQ2 hiring. This is the step where the analysis moves from U.S. analysis and become 
           to become Amazon.com specific.
        7. Provide Amazon.com estimates of permanent, gig, and contract 2029 employees by occupations and 
           geographical level.
        8. Build visualizations to illustrate the differences between both U.S. and Amazon.com employee 
           data in 2019 and 2029 predicted data by geographic locations and occupations for proposed HQ2 
           locations.
           
### E.)	Deliverables:
         
        The deliverables are below:
        a.)	Slide deck with the following:
            1.	an understanding of the project.
            2.	a summary of the original data points.
            3.  classification, clustering, and regression analysis that produce predicted data points.
            4.  estimated 2029 data points alongside 2019 data points with percentage differences.
        b.)	Any programming and data munging code.
        c.)	All data in a cleaned form that was used for analysis. 
        d.)	Trained usable models that can be used to run future data through using another factor 
            (i.e. GDP growth).

            When the project is complete, the project can be listed on Data Science Central and LinkedIn 
            to illustrate competency.

### F.)	Works Cited:

        1.)	Amazon blog, The. Investor Relations. (2018, October 13). Retrieved from https://ir.aboutamazon.com/.
        2.)	"Amazon expanding Detroit presence". Archived from the original on 2016-07-11. 
            Retrieved October 14, 2018.
        3.)	 "Amazon Minneapolis". Archived from the original on 2016-06-10. 
        4.)	Bisnow Team. Forbes. "Amazon HQ2 Shortlist: Details on the 20 Finalists In $5B Sweepstakes." 
            (2018, October 14). Retrieved from 
            https://www.forbes.com/sites/bisnow/2018/01/19/amazon-hq2-shortlist-details-on-the-20-finalists-
            in-5b-sweepstakes/. 
       5.)	Carson, E. (2017). "Amazon will hire 120,000 temporary workers this holiday season." CNET. Retrieved 
            from https://www.cnet.com/news/amazon-plans-to-hire-120000-temporary-workers-this-holiday-season/. 
       6.)	Glassdoor. Amazon Flex Overview. (2018, October 14). Retrieved from 
            https://www.glassdoor.com/Overview/Working-at-Amazon-Flex-EI_IE1324363.11,22.htm.
       7.)	Kim, E. (2017). "Here's who really runs Amazon - and only 2 of the top 38 execs are women." Retrieved 
            from https://www.cnbc.com/2017/11/28/who-are-amazons-top-executives.html. 
       8.)	Kirsner, Scott (December 23, 2011). "Amazon plans Cambridge office". Boston Globe. Archived from the 
            original on January 16, 2013. Retrieved October 14, 2018. 
       9.)	Neibauer, Michael. "Amazon's Herndon employees will earn $114K on average". Archived from the original 
            on June 19, 2015. Retrieved October 14, 2018. 
      10.)	Novak, Shonda (November 12, 2014). "Sources: Amazon.com to bring 200-plus tech jobs to Austin". 
            Austin-American Statesman. Archived from the original on February 26, 2015. Retrieved October 14, 2018. 
      11.)	"Pittsburgh". Archived from the original on January 17, 2017. Retrieved October 14, 2018. 
      12.)	Redman, R. (2018). "Growth spurt for Prime Now grocery delivery." SN. Supermarket News. (2018, October 
            14). 
            Retrieved from https://www.supermarketnews.com/online-retail/growth-spurt-prime-now-grocery-delivery. 
      13.)	Reuters. Fortune. (2018, October 14). Retrieved from 
            http://fortune.com/2016/10/13/amazon-120000-temporary-workers-holiday-season/.
      14.)	"San Luis Obispo". a2z.com. Archived from the original on March 3, 2012. Retrieved October 14, 2018. 
      15.)	Side Husl. Amazon Flex. (2018, October 14). Retrieved from https://sidehusl.com/amazon-flex/. 
